Staged Abstract Interpreters

### 2019-09-08 paper/abstract.tex
Proofread abstract

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter. As an example of \emph{abstraction without regret},
the overhead of these abstraction layers is eliminated in the generated code
after staging. We show that staging abstract interpreters is practical and
useful to optimize static analysis, while requiring less engineering effort,
without compromising soundness. We conduct three case studies, including a
comparison with Boucher and Feeley's abstract compilation, applications to
various control-flow analyses, and a demonstration that can be used for modular
analysis. We also empirically evaluate the effect of staging on the execution
time. The experiment shows an order of magnitude speedup with staging for
control-flow analyses.

>>>>

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstractions and
monadic abstractions. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in one of four modes: as an unstaged
concrete interpreter, a staged concrete interpreter, an unstaged abstract
interpreter, or a staged abstract interpreter. As an example of
\emph{abstraction without regret}, the overhead of these abstraction layers is
eliminated in the generated code after staging. We show that staging abstract
interpreters is practical and useful to optimize static analysis, all while
requiring less engineering effort and without compromising soundness. We conduct
three case studies, including a comparison with Boucher and Feeley's abstract
compilation, applications to various control-flow analyses, and a demonstration
that can be used for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.
